A Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-07 03-semantics.tex
Semantics: major review

\newcommand{\ST}{\1\xrightarrow[~n~]{}\1}
\newcommand{\LL}{\langle}
\newcommand{\RR}{\rangle}
\newcommand{\DS}{\displaystyle}

@@

\begin{figure}
\begin{lstlisting}[mathescape=true]
                                   // primary expressions
  p ::= mem(id)                   (any memory access to `id')
      $|$ await(id)                  (await event `id')
      $|$ emit(id)                   (emit event `id')
      $|$ break                      (loop escape)
                                   // compound expressions
      $|$ if mem(id) then p else p   (conditional)
      $|$ p ; p                      (sequence)
      $|$ loop p                     (repetition)
      $|$ p and p                    (par/and)
      $|$ p or p                     (par/or)
      $|$ fin p                      (finalization)
                                   // derived by semantic rules
      $|$ awaiting(id,n)             (awaiting `id' since sequence number `n')
      $|$ emitting(n)                (emitting on stack level `n')
      $|$ p @ loop p                 (unwinded loop)
      $|$ nop                        (terminated expression)
\end{lstlisting}
\caption{
    Reduced syntax of \CEU.
\label{lst.formal.syntax}
}
\end{figure}

Figure~\ref{lst.formal.syntax} shows the syntax for a subset of \CEU that is 
sufficient to describe all semantic peculiarities of the language.
%
Except for $fin$ and the expressions used internally by the semantics (i.e., 
$awating$, $emitting$, $p~@~loop; p$, and $nop$), all other expressions are 
equivalent to their counterparts in the concrete language.

The $mem(id)$ primitive represents all accesses, assignments, system calls, and 
output events that affect a memory location identified by $id$.
%
According to the synchronous hypothesis of \CEU, $mem$ expressions are 
considered to be atomic and instantaneous.
%
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%
%The special notation $nop$ is used to represent an innocuous $mem$ expression 
%(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
%unused identifier).
%
Note that $mem$ and $await$/$emit$ expressions do not share identifiers, i.e., 
an identifier is either a variable or an event.

@@

The core of our semantics describes how a program reacts to a single external 
input event, i.e., starting from the input event, how the program behaves and 
becomes idle again to proceed to the subsequent reaction.
%
We use a set of small-step operational rules, which are built in such a way 
that at most one transition is possible at any time, resulting in deterministic 
reactions.
%
Each reaction is identified by a ever-increasing $n$ that remains constant 
during the entire reaction.
The transition rules map a program $p$ and a stack of events $S$ in a single 
step to a modified program and stack:
%
\begin{align*}
\LL S, p \RR &\ST
\LL S', p' \RR
    & \textbf{(rule-inner)}
\end{align*}
%
where
%
\begin{align*}
S, S' &\in id^*
    &&(stack~of~event~identifiers: [id_{top}, ..., id_{bottom}]) \\
p, p' &\in P
    && (program~as~described~in~Figure~\ref{lst.formal.syntax}) \\
n     &\in \mathds{N}
    && (unique~identifier~for~the~entire~reaction)
\end{align*}
%
At the beginning of a reaction, the stack is initialized with the occurring 
external event $ext$ ($S=[ext]$), but $emit$ expressions can push new events on 
top of it (we discuss how they are popped further).
The sequence number $n$, which is incremented each reaction, prevents that 
awaiting expressions awake in the same reaction they are reached (the 
\emph{delayed awaits} as explained in Section~\ref{sec.ceu.ints}).

The transition rules for the primary expressions are as follows:
%
{ \setlength{\jot}{20pt}
\begin{align*}
\LL S,\1mem(id) \RR &\ST
\LL S,\1nop \RR
    & \textbf{(mem)}        \\
%%%
\LL S,\1await(id) \RR &\ST
\LL S,\1awaiting(id,n+1) \RR
    & \textbf{(await)}      \\
%%%
\LL id:S,\1awaiting(id,m) \RR &\ST
\LL id:S,\1nop \RR, \2if~m\leq n
    & \textbf{(awake)}   \\
%%%
\LL S,\1emit(id) \RR &\ST
\LL id:S,\1emitting(|S|) \RR
    & \textbf{(emit)}       \\
%%%
\LL S,\1emitting(k) \RR &\ST
\LL S,\1nop \RR, \2 if~k=|S|
    & \textbf{(pop)}        %\\
\end{align*}
}
%
A $mem$ operation executes immediately and becomes a $nop$ to indicate 
termination (rule \textbf{mem}).
An $await$ is transformed into an $awaiting$ (rule \textbf{await}) as an 
artifice to remember the external sequence number $n+1$ it can awake:
an $awaiting$ can only transit to a $nop$ (rule \textbf{awake}) if its referred 
event $id$ matches the top of the stack and it was reached in a previous 
reaction (i.e., sequence number $m \leq n$).
%
%Remember that in \CEU, the \code{await} statement returns the value associated 
%with the corresponding event: the yielded $mem$ represents the operation to 
%query that value.
%
An $emit$ transits to an $emitting$ holding the current stack level ($|S|$ 
stands for the stack length), and pushing the referred event on the stack (rule 
\textbf{emit}).
With the new stack level $|S|+1$, the $emitting(|S|)$ itself cannot transit, as 
rule \textbf{pop} expects its parameter $k$ to match the current stack level.
This trick provides the desired stack-based semantics for internal events.

Proceeding to compound expressions, the rules for conditionals and sequences 
are straightforward:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    { \DS val(id,n) \neq 0 }
%   -----------------------------------------------------------
    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
          \LL S, p \RR }
    & \textbf{(if-true)}       \\
%%%
& \frac
    { \DS val(id,n) = 0 }
%   -----------------------------------------------------------
    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
          \LL S, q \RR }
    & \textbf{(if-false)}       \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~;~q) \RR \ST \LL S', (p'~;~q) \RR }
    & \textbf{(seq-adv)}      \\
%%%
& \LL S, (nop~;~q) \RR \ST  \LL S, q \RR
    & \textbf{(seq-nop)}      \\
%%%
& \LL S, (break~;~q) \RR \ST \LL S, break \RR
    & \textbf{(seq-brk)}
\end{eqnarray*}
}
%
Given that our semantics focuses on control, rules \textbf{if-true} and 
\textbf{if-false} are the only to query $mem$ expressions.
%
The ``magical'' function $val$ receives a memory identifier and the current 
reaction sequence number, returning the current memory value.
%
Although the value here is arbitrary, it is unique in a reaction, because a 
given expression can execute only once within it (remember that $loops$ must 
contain $awaits$ which, from rule \textbf{await}, cannot awake in the same 
reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).

The rules for loops are analogous to sequences, but use \code{`@'} as 
separators to properly bind breaks to their enclosing loops:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL S, (loop~p) \RR \ST \LL S, (p~@~loop~p) \RR
    & \textbf{(loop-expd)}       \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
% -----------------------------------------------------------
    {\DS \LL S, (p~@~loop~q) \RR \ST \LL S', (p'~@~loop~q) \RR }
    & \textbf{(loop-adv)}    \\
%%%
& \LL S, (nop~@~loop~p) \RR \ST \LL S, loop~p \RR
    & \textbf{(loop-nop)}    \\
%%%
& \LL S, (break~@~loop~p) \RR \ST \LL S, nop \RR
    & \textbf{(loop-brk)}
\end{eqnarray*}
}
%
When a program encounters a $loop$, it first expands its body in sequence with 
itself (rule \textbf{loop-expd}).
Rules \textbf{loop-adv} and \textbf{loop-nop} are similar to rules 
\textbf{seq-adv} and \textbf{seq-nop}, advancing the loop until they reach a 
$nop$.
However, what follows the loop is the loop itself (rule \textbf{loop-nop}).
Note that if we used \code{`;'} as a separator in loops, rules 
\textbf{loop-brk} and \textbf{seq-brk} would conflict.
%
Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything into 
a $nop$.
%Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything 
%into a $clear(p)$.
%We cannot simply transform the loop into a $nop$ because its body may be a 
%parallel composition containing finalization blocks.

Proceeding to parallel compositions, the semantic rules for $and$ and $or$ 
always force transitions on their left branches $p$ to occur before their right 
branches $q$:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p'~and~q) \RR }
    & \textbf{(and-adv1)}      \\
%%%
& \frac
    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p~and~q') \RR }
    & \textbf{(and-adv2)}      \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~q) \RR \ST \LL S', (p'~or~q) \RR }
    & \textbf{(or-adv1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
%   -----------------------------------------------------------
    {\DS \LL S (p~or~q) \RR \ST \LL S', (p~or~q') \RR }
    & \textbf{(or-adv2)}   %\\
\end{eqnarray*}
}
%
The deterministic behavior of the semantics relies on the \emph{isBlocked} 
predicate, which is defined in Figure~\ref{fig.isBlocked} and used in rules 
\textbf{and-adv2} and \textbf{or-adv2}.
These rules require the left branch $p$ to be blocked in order to allow the 
right transition from $q$ to $q'$.
%
Basically, the \emph{isBlocked} predicate determines that an expression becomes 
blocked when all of its trails in parallel hang in $awaiting$ and $emitting$ 
expressions that cannot advance.

\begin{figure}
{\small
\begin{align*}
  isBlocked(n,a:S, awaiting(b,m)) &= (a \neq b \1\vee\1 m > n)   \\
  isBlocked(n,S, emitting(s))    &= (|S| \neq s)                     \\
  isBlocked(n,S, (p~;~q))        &= isBlocked(n,S,p)             \\
  isBlocked(n,S, (p~@~loop~q))   &= isBlocked(n,S,p)             \\
  isBlocked(n,S, (p~and~q))      &= isBlocked(n,S,p) \wedge
                                    isBlocked(n,S,q)             \\
  isBlocked(n,S, (p~or~q))       &= isBlocked(n,S,p) \wedge
                                    isBlocked(n,S,q)             \\
  isBlocked(n,S, \_)             &= false \2  (nop,await,      \\
                                  &    \5\5\5\2 emit,break,if,loop)   %\\
\end{align*}
}%
%\rule{14cm}{0.37pt}
\caption{
The recursive predicate $isBlocked$ is true only if all branches in parallel 
are hanged in $awaiting$ or $emitting$ expressions that cannot transit.
\label{fig.isBlocked}
}
\end{figure}

%%%

For a parallel $and$, if one of the sides terminates, the composition is simply 
substituted by the other side (rules \textbf{and-nop1} and \textbf{and-nop2}, 
as follows).
%
For a parallel $or$, if one of the sides terminates, the whole composition 
terminates, also applying the $clear$ function to properly finalize the aborted 
side (rules \textbf{or-nop1} and \textbf{or-nop2}):
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL S, (nop~and~q) \RR \ST \LL S, q \RR
    & \textbf{(and-nop1)}   \\
%%%
& \LL S, (p~and~nop) \RR \ST \LL S, p \RR
    & \textbf{(and-nop2)}   \\
%%%
& \LL S, (nop~or~q) \RR \ST \LL S, clear(q) \RR
    & \textbf{(or-nop1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~nop) \RR \ST \LL S, clear(p) \RR }
    & \textbf{(or-nop2)}   %\\
\end{eqnarray*}
}
%
The $clear$ function, defined in Figure~\ref{fig.formal.clear}, concatenates 
all active $fin$ bodies of the side being aborted, so that they execute before 
the composition rejoins.
Note that there are no transition rules for $fin$ expressions.
This is because once reached, a $fin$ expression halts and will only execute 
when it is aborted by a trail in parallel and is expanded by the $clear$ 
function.
In Section~\ref{sec.formal.fins}, we show how to map a finalization block in 
the concrete language to a $fin$ in the formal semantics.
%
Note also that there is a syntactic restriction that a $fin$ body can only 
contain $mem$ expressions in sequence, i.e., they are guaranteed to execute 
entirely within a reaction.

\begin{figure}[b]
{\small
\begin{align*}
  clear( fin~p )       &= p                   \\
  clear( p~;~q )       &= clear(p)            \\
  clear( p~@~loop~q) ) &= clear(p)            \\
  clear( p~and~q )     &= clear(p)~;~clear(q) \\
  clear( p~or~q )      &= clear(p)~;~clear(q) \\
  clear( \_ )          &= nop
\end{align*}
}%
%\rule{14cm}{0.37pt}
\caption{
The function $clear$ extracts $fin$ expressions in parallel and put their 
bodies in sequence.
\label{fig.formal.clear}
}
\end{figure}

Finally, a $break$ in one of the sides in parallel escapes the closest 
enclosing $loop$, properly aborting the other side by applying the $clear$ 
function:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL S, (break~and~q) \RR \ST \LL S, (clear(q)~;~break) \RR
    & \textbf{(and-brk1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
    & \textbf{(and-brk2)}   \\
%%%
& \LL S, (break~or~q) \ST \LL S, (clear(q)~;~break) \RR
    & \textbf{(or-brk1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
    & \textbf{(or-brk2)}   %\\
\end{eqnarray*}
}
%
A reaction eventually blocks in $awaiting$ and $emitting$ expressions in 
parallel trails.
%
If all trails hangs only in $awaiting$ expressions, it means that the program 
cannot advance in the current reaction.
%
However, $emitting$ expressions are pending in lower stack indexes and should 
eventually resume in the ongoing reaction (see rule \textbf{pop}).
%
Therefore, we define another rule that behaves as \textbf{rule-inner} 
(presented above) if the program can advance, and, otherwise, pops the stack to 
resume the lower level:
%
$$
\frac
    { \DS \LL S,p \RR \ST                   \LL S',p' \RR }
%   -----------------------------------------------------------
    {     \LL S,p \RR \xRightarrow[~~n~~]{} \LL S',p' \RR }
%
\5\5\5
%
\frac
    { \DS isBlocked(n,\1s:S,\1p) }
%   -----------------------------------------------------------
    { \LL s:S,p \RR \xRightarrow[~~n~~]{} \LL S,p \RR }
\5\5\5\textbf{(rule-outer)}
$$
%
To describe a \emph{reaction} in \CEU, i.e., how a program behaves in reaction 
to a single external event, we use the reflexive transitive closure of 
\textbf{rule-outer}:
%
$$
    \LL S,p \RR \xRightarrow[~~n~~]{*} \LL S',p' \RR
$$
%
Finally, to describe the complete execution of a program, we trigger multiple 
``invocations'' of reactions in sequence:
%
\begin{align*}
\LL [e1], p \RR
    & \xRightarrow[~~1~~]{*}
\LL [  ], p' \RR
\\
\LL [e2], p' \RR
    & \xRightarrow[~~2~~]{*}
\LL [  ], p'' \RR
\\
\LL [e3], p'' \RR
    & \xRightarrow[~~3~~]{*}
\LL [  ], p''' \RR
\\
& ...
\end{align*}
%
Each invocation starts with the occurring external event at the top of the 
stack and finishes with a modified program and an empty stack.
After each invocation, we increment the sequence number.

>>>>

\newcommand{\NST}{\1\xrightarrow[nst]\1}
\newcommand{\OUT}{\1\xrightarrow[out]\1}
\newcommand{\LL}{\langle}
\newcommand{\RR}{\rangle}
\newcommand{\DS}{\displaystyle}

@@

\begin{lstlisting}[
  %numbers=left,
  basicstyle=\ttfamily\footnotesize,
  float=h,
  caption={Reduced syntax of \CEU.},
  label={lst.formal.syntax},
  mathescape=true
]
                                   // primary expressions
  p ::= mem(id)                    (any memory access to `id')
      $|$ awaitExt(id)               (await external event `id')
      $|$ awaitInt(id)               (await internal event `id')
      $|$ emitInt(id)                (emit internal event `id')
      $|$ break                      (loop escape)
                                   // compound expressions
      $|$ if mem(id) then p else p   (conditional)
      $|$ p ; p                      (sequence)
      $|$ loop p                     (repetition)
      $|$ p and p                    (par/and)
      $|$ p or p                     (par/or)
      $|$ fin p                      (finalization)
                                   // derived by semantic rules
      $|$ p @loop p                  (unwinded loop)
      $|$ p @and q                   (unwinded par/and)
      $|$ p @or q                    (unwinded par/or)
      $|$ @canrun(n)                 (can run on stack level `n')
      $|$ @nop                       (terminated expression)
\end{lstlisting}

Listing~\ref{lst.formal.syntax} shows the syntax for a subset of \CEU that is 
sufficient to describe all semantic peculiarities of the language.
%
Except for $fin$ and the expressions used internally by the semantics, which
are prefixed with a character $@$, all other expressions are equivalent to
their counterparts in the concrete language.

The $mem(id)$ primitive represents all accesses, assignments, system calls, and
output events that affect a memory location identified by $id$.
%
According to the synchronous hypothesis of \CEU, $mem$ expressions are 
considered to be atomic and instantaneous.
%
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%
%The special notation $nop$ is used to represent an innocuous $mem$ expression 
%(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
%unused identifier).
%
Note that $mem$, $awaitExt$, and $awaitInt$ / $emitInt$ expressions do not
share identifiers, i.e., an identifier is either a variable, an external event,
or an internal event.

@@

The core of our semantics describes how a program reacts to a single external 
input event, i.e., starting from the input event, how the program behaves and 
becomes idle again to proceed to the subsequent reaction.
%
We use a set of small-step operational rules, which are built in such a way 
that at most one transition is possible at any time, resulting in deterministic 
reactions.
%
The transition rules map a triple with a program $p$, a stack level $n$, and an
emitted event $e$ to a modified triple as follows:
%
\begin{align*}
\LL p,  n,  e  \RR &\1\xrightarrow\1
\LL p', n', e' \RR
    & \textbf{(rule-general)}
\end{align*}
%
where
%
\begin{align*}
p, p' &\in P
    && (program~as~described~in~Listing~\ref{lst.formal.syntax})
\\
n, n' &\in \mathds{N}
    && (current~stack~level)
\\
e, e' &\in \mathds{E \cup \{\epsilon\}}
    && (emitted~event,~possibly~none)
\end{align*}
%
At the beginning of a reaction to an input event, the triple is initialized at
stack level $0$ ($n=0$) and with the emitted event $id$ ($e=id$).
%, but \code{emitInt} expressions can increase the stack level.
At the end of a reaction, after a number of transitions, the triple will block
with a (possibly) modified program $p'$, at stack level $0$, with no event
emitted $\epsilon$:
%
\begin{align*}
\LL p,0,ext \RR
    & \1\xrightarrow{*}\1
\LL p',0,\epsilon \RR
\end{align*}
%
We also define two transition rule subtypes $\OUT$ (for \emph{outermost}) and
$\NST$ (for \emph{nested}).

The $\OUT$ rules \textbf{push} and \textbf{pop} are non-recursive definitions
that manipulate the stack level as follows:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    { \DS e \neq \epsilon }
%   -----------------------------------------------------------
    { \DS \LL p,n,e \RR \OUT \LL bcast(p),n+1,\epsilon \RR }
    & \textbf{(push)}   \\
%%%
& \frac
    { \DS n>0, \2 ((p=@nop) \wedge isblocked(n,p)) }
%   -----------------------------------------------------------
    { \DS \LL p,n,\epsilon \RR \OUT \LL p,n-1,\epsilon \RR }
    & \textbf{(pop)}    \\
%%%
%& \LL p,0,\epsilon \RR \1\xrightarrow\1 \bot
    %& \textbf{(end)}    %\\
\end{eqnarray*}
}
%
Rule \textbf{push} matches whenever there is an emitted event in the triple,
and instantly broadcasts the event to the program, which means
    (a) awaking active \code{await} expressions altogether (see $bcast(p)$ in
        Figure~\ref{fig.bcast}),
    (b) creating a nested reaction by increasing the stack level,
    (c) and, at the same time, consuming the event ($e$ becomes $\epsilon$).
%
Note that rule \textbf{push} is the only one in the semantics that matches an
emitted event and also immediately consumes it.
%
The rule \textbf{pop} only decreases the stack level, not affecting the
program, and does not apply if the program is not blocked (see $isblocked(n,p)$
in Figure~\ref{fig.isblocked}).
This condition ensures that an \code{emit} only resumes after its internal
reaction completes, as discussed in Section~\ref{sec.ceu.evts}.

Note that at the beginning of a reaction, an event $ext$ is emitted, which
will trigger rule \textbf{push}, which will immediately raise the stack level
to $1$.
Since rule \textbf{pop} is the only to decrease the stack level and only
applies to a blocked or terminated program, the semantics guarantees that a
reaction only terminates with a blocked program at stack level $0$.

The $\NST$ rules are recursive definitions but do not affect the stack level
and never have an emitted event as a precondition:
%
\begin{align*}
\LL p, n,\epsilon \RR &\NST
\LL p',n,e        \RR
    & \textbf{(rule-inner)}
\end{align*}
%
The distinction between $\OUT$ and $\NST$ prevents rules \textbf{push} and
\textbf{pop} to match and inadvertently modify the current stack level before
completing a nested reaction.
%
A complete reaction behaves as follows:
%
\begin{align*}
a) &\5\5
    \LL p,0,ext \RR
        \1\xrightarrow[out]{push}\1
    \LL q,1,\epsilon \RR
\\
b) &\5\5 \1[ \1\xrightarrow[in]{*}\1
    \LL r,i,e \RR
        \1\xrightarrow[out]\1
    \LL s,j,\epsilon \RR \1]*
\\
c) &\5\5 \1\xrightarrow[in]{*}\1
    \LL t,k,\epsilon \RR
        \1\xrightarrow[out]{pop}\1
    \LL u,0,\epsilon \RR
\end{align*}
%
First, a $\xrightarrow[out]{push}$ starts a nested reaction at level $1$ (case
$a$).
Then, a series of alternations between $\NST$ transitions followed by a single
$\OUT$ transition represent nested reactions and stack operations (case $b$).
Finally, at some point, a $\xrightarrow[out]{pop}$ lowers the stack level to
$0$ and terminates the reaction.

The $\NST$ transition rules for the primary expressions are as follows:
%
{ \setlength{\jot}{20pt}
\begin{align*}
\LL mem(id), n, \epsilon \RR &\NST
\LL @nop, n, \epsilon \RR
    & \textbf{(mem)}        \\
%%%
\LL emit(id), n, \epsilon \RR &\NST
\LL @canrun(n), n, id \RR
    & \textbf{(emitInt)}    \\
%%%
\LL @canrun(n), n, \epsilon \RR &\NST
\LL @nop, n, \epsilon \RR
    & \textbf{(canrun)}     \\
\end{align*}
}
%
A $mem$ operation becomes a $nop$ which indicates termination (rule
\textbf{mem}).
An $emitInt(id)$ generates an event $id$ and transits to a $@canrun(n)$ which
can only resume at level $n$ (rule \textbf{emitInt}).
Since nested rules cannot transit with $e \neq \epsilon$, this rule will cause
rule \textbf{push} to execute at the outer level, creating a new level $n+1$ on
the stack.
Also, with the new stack level, the resulting $@canrun(n)$ itself cannot
transit, providing the desired stack-based semantics for internal events.

Proceeding to compound expressions, the rules for conditionals and sequences 
are straightforward:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    { \DS val(id) \neq 0 }
%   -----------------------------------------------------------
    { \DS \LL (if~mem(id)~then~p~else~q),n,\epsilon \RR \NST
          \LL p, n, \epsilon \RR }
    & \textbf{(if-true)}       \\
%%%
& \frac
    { \DS val(id,n) = 0 }
%   -----------------------------------------------------------
    { \DS \LL (if~mem(id)~then~p~else~q),n,\epsilon \RR \NST
          \LL q,n,\epsilon \RR }
    & \textbf{(if-false)}       \\
%%%
& \frac
    { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
%   -----------------------------------------------------------
    { \DS \LL (p~;~q), n, \epsilon \RR \NST \LL (p'~;~q), n, e \RR }
    & \textbf{(seq-adv)}      \\
%%%
& \LL (@nop~;~q),n,\epsilon \RR \NST  \LL q,n,\epsilon \RR
    & \textbf{(seq-nop)}      \\
%%%
& \LL (break~;~q),n,\epsilon \RR \NST \LL break,n,\epsilon \RR
    & \textbf{(seq-brk)}
\end{eqnarray*}
}
%
Given that our semantics focuses on control, rules \textbf{if-true} and 
\textbf{if-false} are the only to query $mem$ expressions.
%
Function $val$ receives a memory identifier and returns the current memory
value.
%
%Although the value here is arbitrary, it is unique in a reaction, because a 
%given expression can execute only once within it (remember that $loops$ must 
%contain $awaits$ which, from rule \textbf{await}, cannot awake in the same 
%reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).

As determined for nested rules, compound expressions also can only have
$\epsilon$ as a precondition and they never modify $n$.
However, they can still emit an event to nest another reaction.
For instance, in rule \textbf{seq-adv}, if the sub-expression $p$ emits event
$e$, the whole composition also emits $e$.
%However, rules \textbf{push} and \textbf{pop} can only match at the outermost
%level.

The rules for loops are analogous to sequences, but use \code{`@'} as 
separators to properly bind breaks to their enclosing loops:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL (loop~p),n,\epsilon \RR \NST \LL (p~@loop~p), n, \epsilon \RR
    & \textbf{(loop-expd)}       \\
%%%
& \frac
    { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
% -----------------------------------------------------------
    { \DS \LL (p~@loop~q),n,\epsilon \RR \NST \LL (p'~@loop~q), n, e \RR }
    & \textbf{(loop-adv)}    \\
%%%
& \LL (@nop~@loop~p), n, \epsilon \RR \NST \LL (loop~p), n, \epsilon \RR
    & \textbf{(loop-nop)}    \\
%%%
& \LL (break~@loop~p), n, \epsilon \RR \NST \LL @nop, n, \epsilon \RR
    & \textbf{(loop-brk)}
\end{eqnarray*}
}

%
When a program encounters a $loop$, it first expands its body in sequence with 
itself (rule \textbf{loop-expd}).
Rules \textbf{loop-adv} and \textbf{loop-nop} are similar to rules 
\textbf{seq-adv} and \textbf{seq-nop}, advancing the loop until they reach a 
$@nop$.
However, what follows the loop is the loop itself (rule \textbf{loop-nop}).
Note that if we used \code{`;'} as a separator in loops, rules 
\textbf{loop-brk} and \textbf{seq-brk} would conflict.
%
Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything into 
a $@nop$.
%Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything 
%into a $clear(p)$.
%We cannot simply transform the loop into a $nop$ because its body may be a 
%parallel composition containing finalization blocks.

Proceeding to parallel compositions, the semantic rules for $and$ and $or$ 
always force transitions on their left branches $p$ to occur before their right 
branches $q$:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL (p~and~q),n,\epsilon \RR \NST \LL (p~@and~(@canrun(n)~;~q)),n,\epsilon \RR
    & \textbf{(and-expd)}       \\
%%%
& \frac
    { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
%   -----------------------------------------------------------
    { \DS \LL (p~@and~q),n,\epsilon \NST \LL (p'~@and~q),n,e \RR }
    & \textbf{(and-adv1)}      \\
%%%
& \frac
    { \DS isblocked(n,p) \1,\2 \LL q,n,\epsilon \RR \NST \LL q',n,e \RR }
%   -----------------------------------------------------------
    { \DS \LL (p~@and~q),n,\epsilon \RR \NST \LL (p~@and~q'), n, e \RR }
    & \textbf{(and-adv2)}      \\
%%%
& \LL (p~or~q), n, \epsilon \RR \NST \LL (p~@or~(@canrun(n)~;~q)), n, \epsilon \RR
    & \textbf{(or-expd)}       \\
%%%
& \frac
    { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
%   -----------------------------------------------------------
    { \DS \LL (p~@or~q),n,\epsilon \RR \NST \LL (p'~@or~q), n, e \RR }
    & \textbf{(or-adv1)}   \\
%%%
& \frac
    { \DS isblocked(n,p) \1,\2 \LL q,n,\epsilon \RR \NST \LL q',n,e \RR }
%   -----------------------------------------------------------
    { \DS \LL (p~@or~q),n,\epsilon \RR \NST \LL (p~@or~q'), n, e \RR }
    & \textbf{(or-adv2)}   %\\
\end{eqnarray*}
}
%
Rules \textbf{and-expd} and \textbf{or-expd} insert a $@canrun(n)$ at the
beginning of the right branch.
This ensures that an $emitInt$ on the left branch, which would transit to a
$@canrun(n)$, still resumes before the right branch starts.
%
The deterministic behavior of the semantics relies on the \emph{isblocked} 
predicate (Figure~\ref{fig.isblocked}) which appears in rules \textbf{and-adv2}
and \textbf{or-adv2}.
These rules require the left branch $p$ to be blocked in order to allow the 
right branch transition from $q$ to $q'$.

For a parallel $@and$, if one of the sides terminates, the composition is
simply substituted by the other side (rules \textbf{and-nop1} and
\textbf{and-nop2}.
%
For a parallel $or$, if one of the sides terminates, the whole composition 
terminates, also applying the $clear$ function to properly finalize the aborted 
side (rules \textbf{or-nop1} and \textbf{or-nop2}):
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL (@nop~@and~q), n, \epsilon \RR \NST \LL q,n,\epsilon \RR
    & \textbf{(and-nop1)}   \\
%%%
& \LL (p~@and~@nop), n, \epsilon \RR \NST \LL p,n,\epsilon \RR
    & \textbf{(and-nop2)}   \\
%%%
& \LL (@nop~@or~q), n, \epsilon \RR \NST \LL clear(q),n,\epsilon \RR
    & \textbf{(or-nop1)}   \\
%%%
& \frac
    { \DS isblocked(n,p) }
%   -----------------------------------------------------------
    { \DS \LL (p~@or~@nop), n, \epsilon \RR \NST \LL clear(p),n,\epsilon \RR }
    & \textbf{(or-nop2)}   %\\
\end{eqnarray*}
}

%
The $clear$ function (Figure~\ref{fig.formal.clear}) concatenates all active
$fin$ bodies of the side being aborted, so that they execute before the
composition rejoins.
Note that there are no transition rules for $fin$ expressions.
This is because once reached, a $fin$ expression halts and will only execute 
when it is aborted by a trail in parallel and is expanded by the $clear$ 
function.
%In Section~\ref{sec.formal.fins}, we show how to map a finalization block in 
%the concrete language to a $fin$ in the formal semantics.
%
Note also that there is a syntactic restriction that $fin$ bodies cannot
contain awaiting expressions ($awaitExt$, $awaitInt$, and $fin$), i.e., they
are guaranteed to execute entirely within a reaction.

Finally, a $break$ in one of the sides in parallel escapes the closest 
enclosing $loop$, properly aborting the other side by applying the $clear$ 
function:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL (break~@and~q), n, \epsilon \RR \NST \LL (clear(q)~;~break),n,\epsilon \RR
    & \textbf{(and-brk1)}   \\
%%%
& \frac
    { \DS isblocked(n,p) }
%   -----------------------------------------------------------
    { \DS \LL (p~@and~break), n, \epsilon \RR \NST \LL (clear(p)~;~break),n,\epsilon \RR }
    & \textbf{(and-brk2)}   \\
%%%
& \LL (break~@or~q),n,\epsilon \RR \NST \LL (clear(q)~;~break),n,\epsilon \RR
    & \textbf{(or-brk1)}   \\
%%%
& \frac
    { \DS isblocked(n,p) }
%   -----------------------------------------------------------
    { \DS \LL (p~@or~break),n,\epsilon \RR \NST \LL (clear(p)~;~break),n,\epsilon \RR }
    & \textbf{(or-brk2)}   %\\
\end{eqnarray*}
}
%
A reaction eventually blocks in $awaitExt$, $awaitInt$, $fin$, and $@canrun$
expressions in parallel trails.
%
If no trails are blocked in $@canrun$ expressions, it means that the program 
cannot advance in the current reaction.
%
However, $@canrun$ expressions can still resume in lower stack indexes and will
eventually resume in the current reaction (see rule \textbf{pop}).
